
% Thesis Abstract -----------------------------------------------------


%\begin{abstractslong}    %uncommenting this line, gives a different abstract heading
\begin{abstracts}        %this creates the heading for the abstract page

The structure of DNA, chromosomes and genome organization is a topic that has
fascinated the field of biology for many years. Most research focused on the
one-dimensional structure of the genome, studying the linear organizations of
genes and genomes and their link with gene expression and regulation,
splicing, DNA methylation, \dots. Yet, spatial and temporal three-dimensional
(3D) genome architecture is also thought to play an important role in many
genomic functions.

Chromosome conformation capture (3C) based methods, coupled with next
generation sequencing (NGS), allow the measurement, in a single experiment, of
genome wide physical interactions between pairs of loci, thus enabling to
unravel the secrets behind 3D organization of genomes. These new technologies
have paved the way towards a systematic and genome wide analysis of how DNA
folds into the nucleus and opened new avenues to understanding many biological
process, such as gene regulation, DNA replication and repair, somatic copy
number alterations and epigenetic changes. Yet, 3C technologies, as any new
biotechnology, now poses important computational and theoretical challenges
for which mathematically well grounded methods need to be developped.

In this thesis, we attempt to address some of the challenges faced while
analysing such data.

The first chapter is dedicated to developping a robust and accurate method to
infer a 3D model of the genome from Hi-C data. Previous methods often
formulated the inference as an optimization problem akin to {\em
multidimensional scaling } (MDS) based on an {\em ad hoc} conversion of
contact counts into euclidean {\em wish distances}. Chromosomes are modeled
with a beads on a string model, and the methods attempt to place the beads in
a 3D euclidean space to fullfill an number of, often non convex, constraints
and such that the pairwise distances between beads are as close as possible
to the corresponding {\em wish distances}. These approaches rely on dubious
hypothesis to convert contact counts into {\em wish distances}, challenging
the accuracy of the final 3D model. Another limitation is the MDS formulation
which is only intuitively motivated, and not grounded on a clear statistical
model. To alleviate these problems, our method models contact counts as a
Poisson distribution where the
intensity is proportional to the spatial distance between elements
interacting. We then formulate the 3D structure inference as a maximum
likelihood problem. We demonstrate that our method infers robust and stable
models across resolutions and datasets.

The second chapter focuses on the genome architecture of the {\em P.
falciparum}, a small parasite responsible for the deadliest and most virulent
form of human malaria. This project was biologically driven and aimed at
understanding whether and how the 3D structure of the genome related to gene
expression and regulation at different time points in the complex life cycle
of the parasite. In collaboration with Le Roch lab and Noble lab, we built 3D
models of the genome at three time points which resulted in a complex genome
architecture indicative of a strong association between the spatial genome and
gene expression.

The last chapter tackles a very different question, also based on 3C-based
data. Initially developped to probe the 3D architecture of the chromosomes,
Hi-C and related techniques have recently been re-purposed for diverse
applications: \textit{de novo} genome assembly, deconvolution of metagenomic
samples and genome annotations. We describe in this chapter a novel method,
Centurion, that jointly infers the locations of all centromeres in a single
genome from Hi-C data, using the centromeres' tendency to strongly colocalize
in the nucleus.  Indeed, centromeres are essential for proper chromosome
segregation, yet, despite extensive research, centromere locations are unknown
for many yeast species. We demonstrate the robustness of our approach on
datasets with low and high coverage on well annotated organisms. We then
predict centromere coordinates for 6 yeast species that currently lack those
annotations.

During the course of my phd, I have collaborated on several other projects,
for which my contribution were minor and thus which I will not describe in the
main part of this manuscript. The corresponding papers can be found in
appendix. The first project consists in the development of a complete pipeline
to preprocess Hi-C data from reads to normalized contact counts. I have worked
on a fast and memory efficient python implementation of the normalization.
Despite its simplicity, this implementation is to our
knowledge the fastest implementation existing so far. The second paper is a
review of the epigenetics of the {\em P. falciparum} following our first paper
on the 3D structure of this parasite. The last project extends the Hi-C
protocol to detect interaction between triplets and quadruplets of loci in
addition to the usual pairwise interaction. My contribution to this last paper
is the development of a method to infer the 3D structure of polyploid method
which we applied to the KBM7 nearly haploid human cell line.


\end{abstracts}
%\end{abstractlongs}


% ---------------------------------------------------------------------- 
